The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-08 kalman_walk_paper.tex
changed title, fixed figure six Y-axis label and caption

Thus far we have seen that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
This begs the question: does evolution shift molecular mechanisms
fast enough to be a significant driver of speciation?
To approach this question,
we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems
due to the action of recombination, mutation, and demographic noise.
Although this is motivated by the results on linear systems above,
the quantitative genetics calculations are more general,
and only depend on the presence of genetic variation and a continuous set of phenotypically equivalent systems.

@@

\paragraph{The origin of species not by means of natural selection?}
% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
% other mechanisms also probly important
% but this one may be strong since happens in all networks at once
Although, as classically formulated, 
the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch2000origin} 
and under stabilizing selection \citep{fierst},
previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
(Also note that the mechanism underlying hybrid inviability here 
resembles the Dobzhansky-Muller model,
but is more similar to the ``pathway model'' of \citet{lindtke2015genetic}.)
Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual}.
In contrast, the model we develop here suggests that even under strictly neutral conditions, 
system drift can lead to speciation, at a rate fast enough 
to play a substantial role in species formation across the tree of life. 
Our results show that hybrids, under neutral processes, break down as a function of genetic distance,
and so may, in part, explain broad patterns
such as the relationship between molecular divergence and genetic isolation 
% across diverse taxa and ecologies 
seen by \citet{roux2016shedding}, 
and the clocklike speciation patterns observed by \citet{hedges2015tree}.
% (as diversification is not observed to be accelerated following mass extinctions).

>>>>

Thus far we have seen that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
This leads to the question: does evolution shift molecular mechanisms
fast enough to be a significant driver of speciation?
To approach this question,
we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems
due to the action of recombination, mutation, and demographic noise.
Although this is motivated by the results on linear systems above,
the quantitative genetics calculations are more general,
and only depend on the presence of genetic variation and a continuous set of phenotypically equivalent systems.

@@

\paragraph{On the origin of species not by means of natural selection?}
% this is neutral, in contrast to other models. although some folks advocate nonadaptive speciation.
% other mechanisms also probly important
% but this one may be strong since happens in all networks at once
Although, as classically formulated, 
the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary},
and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch2000origin} 
and under stabilizing selection \citep{fierst},
previous authors have argued that neutral processes are likely too slow to be a significant driver of speciation \citep{nei1983models,seehausen2014genomics}.
(Also note that the mechanism underlying hybrid inviability here 
resembles the Dobzhansky-Muller model,
but is more similar to the ``pathway model'' of \citet{lindtke2015genetic}.)
Using simulations, \citet{porter2002speciation} demonstrated the accumulation of hybrid incompatibilities under directional, but not stabilizing selection, and \citet{palmer2009dynamics} observed the appearance of incompatibilities in populations evolving in a constant environment, only if the populations were suboptimally adapted.
This, in light of the few known incompatibilities, has lead some to conclude that hybrid incompatibility is typically a byproduct of positive selection \citep{orr2004speciation, schluter2009evidence} 
or a consequence of genetic conflict \citep{presgraves2010molecular, crespi2013conflictual}.
In contrast, the model we develop here suggests that even under strictly neutral conditions, 
system drift can lead to speciation, at a rate fast enough 
to play a substantial role in species formation across the tree of life. 
Our results show that hybrids, under neutral processes, break down as a function of genetic distance,
and so may, in part, explain broad patterns
such as the relationship between molecular divergence and genetic isolation 
% across diverse taxa and ecologies 
seen by \citet{roux2016shedding}, 
and the clocklike speciation patterns observed by \citet{hedges2015tree}.
% (as diversification is not observed to be accelerated following mass extinctions).
